---
title: "Correlation_SFS"
author: "Charlotte Bunnenberg"
date: "23 3 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries ----
library(PhaseTypeR)
library(tidyverse)
library(grid)
library(gridExtra)
library(reshape2)
# Plot graph
library(igraph)
library(ggplot2)
library("ggpubr")
library(patchwork)
# ggplot settings
theme_set(
  theme_minimal() +
    theme(legend.position = "right")
  )
cbp1 <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
source("Functions.R")
```

# Figure 4.1: 
Correlation between the number of segregating sites at two loci and the recombination rate ρ, for different mutation rates θ = (0.1, 1, 10) and sample sizes n = (2, 4, 8).
```{r}
# Initialization
# Create data frame
df <- data.frame(matrix(ncol = 4, nrow = 0))
colnames(df) <- c('Recombination.Rate', 'Correlation', 'm', "n")

n_list <-c(2,4,8)
m_list <- c(0.1,1,10)

for (n in n_list){
  for (m in m_list){
    for(rho in seq(0, 6, by=0.5)){
      #generate rewards, S and a
      Output <- IntensityMatrix(n, rho)
      # create MPH
      MPH_gen <- MPH(Output[["subIntMat"]], Output[["initProbs"]], Output[["rewardMatMRCA"]])
      #append df with rho, cov and type
      
      df[nrow(df) + 1,] = c(rho, corr_SFS(MPH_gen, m),m,n)
    }
  }
}

# Create plot
plt_cor <- ggplot(data = df, aes(x=Recombination.Rate, y=Correlation, color= factor(m), group=interaction(m, n))) + geom_line(aes(linetype=factor(n)))+ labs(colour =  expression(theta), linetype = "Sample size")+scale_colour_manual(values=c("#E69F00",  "#009E73","#56B4E9"))+xlab(expression(rho))+
  scale_linetype_manual(values=c("solid","longdash","dotted"))
#ggsave(file="cor_rho_SFS.pdf", plot=plt_cor, width=6, height=3)
```

###################################################################
# Estimate recombination rate using method of moments procedure

Call following to estimate the recombination rate for a sample, with
- data_joint = the given data set
- m_estimate = the estimated mutation rate
- n = the sample size
```{r}
optim(par=1, fn=MoM_rho_optim, n = n, m_estimate = m_estimate, data_joint = data, lower = 0, upper = 100, method = 'L-BFGS-B')
```

# Function to generate plot (like figure 4.2) for a given recombination rate, sample size and mutation rate
```{r}
plot_rho_dist_MoM <- function(rho,n,m){
  # Initialize
  Output <- IntensityMatrix(n,rho)
  MPH_gen <- MPH(Output[["subIntMat"]], Output[["initProbs"]], Output[["rewardMatTotal"]])
  optim_MoM <- vector()
  
  for (i in seq(1,100,1)){
  # Simulate data: poisson distributed number of mutations (use rpois-function)
  # on tree of random height (use rMPH-function)     
    nsim = 1000
    sim_MPH = rMPH(nsim, MPH_gen)
    sim_pois_1 = Vectorize(rpois, 'lambda')(1, m*sim_MPH[,1])
    sim_pois_2 = Vectorize(rpois, 'lambda')(1, m*sim_MPH[,2])
    data_joint <- cbind(sim_pois_1,sim_pois_2)
    
    # If wanted an the mutation rate can be estimated:
    # m_estimate <- (watterson(n,c(sim_pois_1, sim_pois_2)))/2
    # If instead using the actual mutation rate:
    m_estimate <- m
    
    # Estimate recombination rate
    optim_MoM[length(optim_MoM)+1] <- optim(par=1, fn=MoM_rho_optim, n = n, m_estimate = m_estimate, data_joint = data_joint, lower = 0, upper = 100, method = 'L-BFGS-B')$par
  }
  
  # Generate ggplot
  plt <- data_frame(estimated_rho = optim_MoM) %>%
            ggplot(aes(x = estimated_rho )) + 
                  geom_histogram(aes(y=..density..), colour="#999999", fill="white")+
   geom_density(alpha=.2, color = "#56B4E9",fill="#56B4E9") +
    geom_vline(aes(xintercept=mean(estimated_rho), color = "Simulation mean"),
              linetype="dashed", size=0.5)+
    geom_vline(aes(xintercept=rho, color="True"),size=0.5)+
    scale_color_manual(name = expression(rho), values = c("Simulation mean" = "#56B4E9", "True" = "#E69F00"))+
    xlab(expression(rho))+ ylab("Density")+
    ggtitle(label= paste("Approximation error: ", round(abs(rho - mean(optim_MoM)),5)))+
    theme(plot.title = element_text(size = 8), axis.title.x =element_text(size=9), axis.title.y =element_text(size=9) )

  return(plt)
}
```

# Figure 4.2: 
Estimated recombination rate, using the method of moments, for 1.000 samples for a 100 iterations, represented by bars, with true recombination rate ρ = 1 sample size n = 2 and mutation rate θ = 1. The density of the estimates is shown in blue. The orange line represents the true recombination rate, while the dashed blue line represents the mean of the estimated recombination rates.
```{r}
plt_1 <- plot_rho_dist_MoM(1,2,1)
plt_1
```

# Figure 4.3: 
Estimated recombination rate and true recombination rate,
for 0 < ρ < 15, sample size n = 2 and two mutation rates θ = 0.1
(orange) and θ = 10 (blue). The grey line is an orientation and refers
to the true ρ on both axes.
```{r}
#Initialize 
n <- 2
m <- 0.1
m_high <- 10
compare_df_mom_100sim_1000sample <- data.frame(rho=double(),
                 real_corr=double(),
                 sim_rho=double(),
                 sim_rho_high_m = double())
for(r in seq(0,15,0.25)){
  Output <- IntensityMatrix(n,r)
  MPH_gen <- MPH(Output[["subIntMat"]], Output[["initProbs"]], Output[["rewardMatTotal"]])
  optim_MoM <- vector()
  optim_MoM_high <- vector()
# Simulate data: poisson distributed number of mutations (use rpois-function) on tree of random height (use rMPH-function) 
  for (i in seq(1,100,1)){
    nsim = 1000
    sim_MPH = rMPH(nsim, MPH_gen)
    # For θ = 0.1:
    sim_pois_1 = Vectorize(rpois, 'lambda')(1, m*sim_MPH[,1])
    sim_pois_2 = Vectorize(rpois, 'lambda')(1, m*sim_MPH[,2])
    data_joint <- cbind(sim_pois_1,sim_pois_2)
    # Estimate recombination rate 
    optim_MoM[length(optim_MoM)+1] <- optim(par=1, fn=MoM_rho_optim, n = n, m_estimate = m, data_joint = data_joint, lower = 0, upper = 100, method = 'L-BFGS-B')$par 
    
    # For θ = 10:
    sim_pois_1_high = Vectorize(rpois, 'lambda')(1, m_high*sim_MPH[,1])
    sim_pois_2_high = Vectorize(rpois, 'lambda')(1, m_high*sim_MPH[,2])
    data_joint_high <- cbind(sim_pois_1_high,sim_pois_2_high)
    # Estimate recombination rate 
    optim_MoM_high[length(optim_MoM_high)+1] <- optim(par=1, fn=MoM_rho_optim, n = n, m_estimate = m_high, data_joint = data_joint_high, lower = 0, upper = 100, method = 'L-BFGS-B')$par 
  } 
  # Calculate mean of estimated recombination rate over the 100 simulations
  real_corr_r <- corr_SFS(MPH_gen, m)
  sim_rho <- mean(optim_MoM)
  sim_rho_high_m <- mean(optim_MoM_high)
  compare_df_mom_100sim_1000sample[nrow(compare_df_mom_100sim_1000sample)+1,]<- c(r, real_corr_r,sim_rho,sim_rho_high_m)
}
# Plot results
plt <- compare_df_mom_100sim_1000sample %>% ggplot(aes(x = rho )) + 
  geom_abline(aes(slope = 1, intercept = 0), colour="#999999") +
    geom_point(stat="identity",aes(y=sim_rho), colour="#E69F00", size = 0.8)+
    geom_point(stat="identity",aes(y=sim_rho_high_m), colour="#56B4E9", size = 0.8)+
    ylab( expression(rho~~"Estimate"))+ xlab(expression(rho))+
    theme(axis.title.x =element_text(size=9), axis.title.y =element_text(size=9) )
#ggsave(file="cor_estimateRho_MoM_m01_m10_50_500.pdf", plot=plt, width=6, height=3)

```

###################################################################
# Estimate recombination rate using maximum likelihood procedure

Call following to estimate the recombination rate for a sample, with
- data_joint = the given data set
- m_estimate = the estimated mutation rate
- n = the sample size
```{r}
optim(par=0.01, fn=ML_prob, n = n, m_estimate = m_estimate, data_joint = data, lower = 0.001, upper = 100, method = 'L-BFGS-B')
```

# Function to generate plot (like figure 4.4) for a given recombination rate, sample size and mutation rate
```{r}
# function to produce plot
plot_rho_dist_prob <- function(rho,n,m){
  Output <- IntensityMatrix(n,rho)
  MPH_gen <- MPH(Output[["subIntMat"]], Output[["initProbs"]], Output[["rewardMatTotal"]])
  rho_est_list_prob <- vector()

  for (i in seq(0,100,1)){
  # Simulate data (sample size = 1000)
    nsim = 1000
    sim_MPH = rMPH(nsim, MPH_gen)
    sim_pois_1 = Vectorize(rpois, 'lambda')(1, m*sim_MPH[,1])
    sim_pois_2 = Vectorize(rpois, 'lambda')(1, m*sim_MPH[,2])
    data_joint <- data.frame(cbind(sim_pois_1,sim_pois_2))

    # If wanted an the mutation rate can be estimated:
    # m_estimate <- (watterson(n,c(sim_pois_1, sim_pois_2)))/2
    # If instead using the actual mutation rate:
    m_estimate <- m
    
    # Estimate recombination rate
    optim_output <- optim(par=0.01, fn=ML_prob, n = n, m_estimate = m_estimate, data_joint = data_joint, lower = 0.001, upper = 100, method = 'L-BFGS-B')
    # Exclude the boundaries
    if(optim_output$par < 100 & optim_output$par > 0.0001){
          rho_est_list_prob[length(rho_est_list_prob)+1] <- optim_output$par
    }
  }
# Create ggplot 
plt <- data_frame(estimated_rho = rho_est_list_prob) %>%
            ggplot(aes(x = estimated_rho )) + 
                  geom_histogram(aes(y=..density..), colour="#999999", fill="white")+
   geom_density(alpha=.2, color = "#56B4E9",fill="#56B4E9") +
    geom_vline(aes(xintercept=mean(estimated_rho), color = "Simulation mean"),
              linetype="dashed", size=0.5)+
    geom_vline(aes(xintercept=rho, color="True"),size=0.5)+
    scale_color_manual(name = expression(rho), values = c("Simulation mean" = "#56B4E9", "True" = "#E69F00"))+
    xlab(expression(rho))+ ylab("Density")+ 
    theme(plot.title = element_text(size = 8), axis.title.x =element_text(size=9), axis.title.y =element_text(size=9) )+
    ggtitle(label= paste("Average Approximation error: ", round(abs(rho - mean(rho_est_list_prob)),5)))
return(plt)
}
```

# Figure 4.4: 
Estimated recombination rate, using the maximum likelihood, for 1.000 samples for a 100 iterations, represented by bars, with true recombination rate ρ = 1 sample size n = 2 and mutation rate θ = 0.1.
The density of the estimates is shown in blue. The orange line represents the true recombination rate, while the dashed blue line represents the mean of the estimated recombination rates.
```{r}
figure4_4 <- plot_rho_dist_prob(1,2,0.1)
figure4_4
```

# Figure 4.5: Estimated recombination rate and true recombination rate,
for 0 < ρ < 15, sample size n = 2 and two mutation rates θ = 0.1
(orange) and θ = 1 (green).
```{r}
n <- 2
m <- 0.1
m_high <- 1
compare_df_ml_100sim_1000sample <- data.frame(
                 sim_rho=double(),
                 sim_rho_high_m = double())
for(r in seq(0,15,0.25)){
  Output <- IntensityMatrix(n,r)
  MPH_gen <- MPH(Output[["subIntMat"]], Output[["initProbs"]], Output[["rewardMatTotal"]])
  optim_ml <- vector()
  optim_ml_high <- vector()
  for (i in seq(1,100,1)){
    #simulate data: pois dist # of mutations on tree of random height rMPH
    nsim = 10
    sim_MPH = rMPH(nsim, MPH_gen)
    sim_pois_1 = Vectorize(rpois, 'lambda')(1, m*sim_MPH[,1])
    sim_pois_2 = Vectorize(rpois, 'lambda')(1, m*sim_MPH[,2])
    data_joint <- cbind(sim_pois_1,sim_pois_2)
    optim_ml[length(optim_ml)+1] <- optim(par=0.00001, fn=ML_prob, n = n, m_estimate = m, data_joint = data_joint, lower = 0.0001, upper = 100, method = 'L-BFGS-B')$par
  #high mutation rate 
    sim_pois_1_high = Vectorize(rpois, 'lambda')(1, m_high*sim_MPH[,1])
    sim_pois_2_high = Vectorize(rpois, 'lambda')(1, m_high*sim_MPH[,2])
    data_joint_high <- cbind(sim_pois_1_high,sim_pois_2_high)
    optim_ml_high[length(optim_ml_high)+1] <- optim(par=0.00001, fn=ML_prob, n = n, m_estimate = m_high, data_joint = data_joint_high, lower = 0.0001, upper = 100, method = 'L-BFGS-B')$par
  } 
  sim_rho <- mean(optim_MoM)
  sim_rho_high_m <- mean(optim_MoM_high)
  compare_df_ml_100sim_1000sample[nrow(compare_df_ml_100sim_1000sample)+1,]<- c(sim_rho,sim_rho_high_m)
}
#compare_df_1_10 <- compare_df
plt <- compare_df_ml_100sim_1000sample %>% ggplot(aes(x = rho )) + 
  geom_abline(aes(slope = 1, intercept = 0), colour="#999999") +
    geom_point(stat="identity",aes(y=sim_rho), colour="#E69F00", size = 0.8)+
    geom_point(stat="identity",aes(y=sim_rho_high_m), colour="#56B4E9", size = 0.8)+
    ylab( expression(rho~~"Estimate"))+ xlab(expression(rho))+
    theme(axis.title.x =element_text(size=9), axis.title.y =element_text(size=9) )
#ggsave(file="cor_estimateRho_ML_m01_m1_50_500.pdf", plot=plt, width=6, height=3)

```

################################################################
# Compare the two recombination rate estimation methods

# Figure 4.6: Estimated recombination rates for ρ = {0, ..., 5}, using the maximum likelihood procedure (◦) and the method of moments procedure (×). The estimates are based on the mean over 100 samples, each of size 1.000. With sample size n = 2 and mutation rate θ = 0.1.
```{r}
# Initialize
n <- 2
m <- 0.1
compare_methods_line <- data.frame(rho=double(),
                 sim_rho_ML=double(),
                 sim_rho_MoM = double())
for(rho in seq(0,5,1)){
  Output <- IntensityMatrix(n,rho)
  MPH_gen <- MPH(Output[["subIntMat"]], Output[["initProbs"]], Output[["rewardMatTotal"]])
  rho_est_list_prob <- vector()
  rho_est_list_mom <- vector()
  
  for (i in seq(1,100,1)){
    # Simulate data
    nsim = 1000
    sim_MPH = rMPH(nsim, MPH_gen)
    sim_pois_1 = Vectorize(rpois, 'lambda')(1, m*sim_MPH[,1])
    sim_pois_2 = Vectorize(rpois, 'lambda')(1, m*sim_MPH[,2])
    data_joint <- cbind(sim_pois_1,sim_pois_2)
    
    # Maximum likelihood estimate
    Ml_par <- optim(par=1, fn=ML_prob, n =n, m_estimate = m, data_joint = data_joint, lower = 0.0001, upper = 100, method = 'L-BFGS-B')$par
      rho_est_list_prob[length(rho_est_list_prob)+1] <- Ml_par
    # Method of moments estimate
    MoM_par <- optim(par=1, fn=MoM_rho_optim, n = n, m_estimate = m, data_joint = data_joint, lower = 0.0001, upper = 100, method = 'L-BFGS-B')$par
      rho_est_list_mom[length(rho_est_list_mom)+1] <-MoM_par
  }
  compare_methods_line[nrow(compare_methods_line)+1,]<- c(rho, mean(rho_est_list_prob), mean(rho_est_list_mom))
}
plot_estimates <- ggplot(compare_methods_line,aes(x = rho))+ geom_point(aes(y = sim_rho_ML),colour = "#E69F00", shape = 1) + geom_point(aes(y = sim_rho_MoM),colour = "#E69F00", shape = 3)+ geom_line(aes(y = rho), colour = "#999999")+xlab(expression(rho))+ylab(expression(rho~~"Estimate"))
#ggsave(file="compare_methods_accuracy.pdf", plot=plot_estimates, width=6, height=3)
```

# Figure 4.7: Runtime of the two recombination rate estimation methods maximum likelihood (◦) and method of moments (×) for different mutation rates θ. The runtime refers to the time spent within the estimation method when estimating the recombination rate for a data set of size 1.000. With each sample of size n = 2 and mutation rate ρ = 1.
```{r}
# Initialize
n <- 2
rho <- 1
compare_methods_line <- data.frame(mutation_rate=double(),
                 sim_rho_ML=double(),
                 sim_rho_MoM = double(),
                mean_runtime_ML = double(),
                mean_runtime_MoM = double(),
                total_runtime_ML = double(),
                total_runtime_MoM = double())
Output <- IntensityMatrix(n,rho)
MPH_gen <- MPH(Output[["subIntMat"]], Output[["initProbs"]], Output[["rewardMatTotal"]])
for(m in seq(0.1,1,0.1)){
  rho_est_list_prob <- vector()
  rho_est_list_mom <- vector()
  rho_est_list_MoM_runtime <- vector()
  rho_est_list_ML_runtime <- vector()
  for (i in seq(1,100,1)){
    # Simulate data
    nsim = 1000
    sim_MPH = rMPH(nsim, MPH_gen)
    sim_pois_1 = Vectorize(rpois, 'lambda')(1, m*sim_MPH[,1])
    sim_pois_2 = Vectorize(rpois, 'lambda')(1, m*sim_MPH[,2])
    data_joint <- cbind(sim_pois_1,sim_pois_2)
    
    # Measure runtime by using Sys.time()-function
    # Maximum likelihood estimation
    start_time_ML <- Sys.time()
    Ml_par <- optim(par=1, fn=ML_prob, n =n, m_estimate = m, data_joint = data_joint, lower = 0.0001, upper = 100, method = 'L-BFGS-B')$par
    end_time_ML <- Sys.time()
      rho_est_list_ML_runtime[length(rho_est_list_ML_runtime)+1] <- end_time_ML -start_time_ML
      rho_est_list_prob[length(rho_est_list_prob)+1] <- Ml_par
    
    # Method of moments estimation
    start_time_MoM <- Sys.time()
    MoM_par <- optim(par=1, fn=MoM_rho_optim, n = n, m_estimate = m, data_joint = data_joint, lower = 0.0001, upper = 100, method = 'L-BFGS-B')$par
    end_time_MoM <- Sys.time()
      rho_est_list_mom[length(rho_est_list_mom)+1] <-MoM_par
      rho_est_list_MoM_runtime[length(rho_est_list_MoM_runtime)+1] <- end_time_MoM -start_time_MoM
  }
  compare_methods_line[nrow(compare_methods_line)+1,]<- c(m, mean(rho_est_list_prob), mean(rho_est_list_mom), mean(rho_est_list_ML_runtime), mean(rho_est_list_MoM_runtime), sum(rho_est_list_ML_runtime), sum(rho_est_list_MoM_runtime))
}

# Plot
plot_runtime <- ggplot(compare_methods_line,aes(x = mutation_rate))+ geom_point(aes(y = total_runtime_ML),colour = "#999999", shape = 1) + geom_point(aes(y = total_runtime_MoM),colour = "#999999", shape = 3)+xlab(expression(theta))+ylab("runtime [ms]")

#ggsave(file="compare_runtime.pdf", plot=plot_runtime, width=6, height=3)
```
