---
title: "TestFunctions"
author: "Charlotte Bunnenberg"
date: "21 3 2023"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
library(PhaseTypeR)
library(tidyverse)
library(grid)
library(gridExtra)
library(reshape2)
#plot graph
library(igraph)
library(ggplot2)
# ggplot settings
theme_set(
  theme_minimal() +
    theme(legend.position = "right")
  )
cbp1 <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
source("Functions.R")
```

# CHECK: Reduce-state-space function
```{r}
Output <- IntensityMatrix(2,2)
Output_reduced <- reduce_S(Output)

# create MPH
MPH_gen <- MPH(Output[["subIntMat"]], Output[["initProbs"]], Output[["rewardMatTotal"]])
MPH_gen_reduced <- MPH(Output_reduced[["subIntMat"]], Output_reduced[["initProbs"]], Output_reduced[["rewardMatTotal"]])
# Plot Density function
x <- seq(0, 10, 0.1)
#plot rho = 0
y <- dMPH(x, MPH_gen)
y1 <- dMPH(x, MPH_gen_reduced)
plot(x, y[1,], type = 'l', col = 'orange')
lines(x, y1[1,],type = 'p', col = 'blue')
legend(x = "topright", text.font = 4,legend=c(expression('S normal, T'[total]), expression('S reduced, T'[total])),
       col=c("orange", "blue"), lty=1)
title('Density function of left tree height comparing normal to reduced S\n starting at (2,0,0), n=2, rho=2')

# starting at (0,2,1)
init_p <- c(0,0,0,0,0,1,0)
init_p_reduced <- c(0,0,0,1,0)
# create MPH
MPH_gen <- MPH(Output[["subIntMat"]], init_p, Output[["rewardMatTotal"]])
MPH_gen_reduced <- MPH(Output_reduced[["subIntMat"]],init_p_reduced, Output_reduced[["rewardMatTotal"]])
# Plot Density function
x <- seq(0, 10, 0.1)
#plot rho = 0
y <- dMPH(x, MPH_gen)
y1 <- dMPH(x, MPH_gen_reduced)
plot(x, y[1,], type = 'l', col = 'orange')
lines(x, y1[1,],type = 'p', col = 'blue')
legend(x = "topright", text.font = 4,legend=c(expression('S normal, T'[total]), expression('S reduced, T'[total])),
       col=c("orange", "blue"), lty=1)
title('Density function of left tree height comparing normal to reduced S\n starting at (0,2,1), n=2, rho=2')
```

# CHECK: Compare the attributes correlation and covariance of the generated phase-type objects to the expected results from Wakeley's book for different recombination rates and sample size n = 2.

```{r}
# Generate phase-type object for rho = {0,4,10}
Output0 <- IntensityMatrix(2, 0)
Output4 <- IntensityMatrix(2,4)
Output10 <- IntensityMatrix(2,10)

MPH_gen0 <- MPH(Output0[["subIntMat"]], Output0[["initProbs"]], Output0[["rewardMatMRCA"]])
MPH_gen4 <- MPH(Output4[["subIntMat"]], Output4[["initProbs"]], Output4[["rewardMatMRCA"]])
MPH_gen10 <- MPH(Output10[["subIntMat"]], Output10[["initProbs"]], Output10[["rewardMatMRCA"]])

x <- seq(0, 10000, 1)
# Generate random tree length data sets
rho0 <- cbind(rMPH(x, MPH_gen0),0)
rho4 <- cbind(rMPH(x, MPH_gen4),4)
rho10 <- cbind(rMPH(x, MPH_gen10),10)

# Print the observed and expected covariance and correlation for the given objects:

#rho = 0: covariance and correlation should be 1
print("rho = 0")
print("correlation == covariance of tree length: (should be 1)")
cor(rho0[,1], rho0[,2])
cov_1(0) 

print("covariance: (should be 1)")
cov(rho0[,1], rho0[,2]) 
cov_1(0)

print("variance:")
var(rho0[,1])
var(MPH_gen0)

print("mean: (should be 1)")
mean(rho0[,2])
mean(MPH_gen0)

# rho = 4
print("rho = 4")
print("correlation:")
cor(rho4[,1], rho4[,2])
cov_1(4)
print("variance:")
var(rho4[,1])
var(MPH_gen4, 1)
print("covariance:")
cov(x = rho4[,1], y = rho4[,2])
cov_1(4)
print("mean:")
mean(rho4[,2])
mean(MPH_gen4)

# rho = 10
print("rho = 10")
print("correlation:")
cor(rho10[,1], rho10[,2])
cov_1(10)
print("variance:")
var(rho10[,1])
var(MPH_gen10, 1)
print("covariance:")
cov(x = rho10[,1], y = rho10[,2])
cov_1(10)
print("mean:")
mean(rho10[,2])
mean(MPH_gen10)
```


# Plot: Relation between tree heights of the two loci for different rho,
  using the objects generated above!
  Observation: Density-curves are the same, but correlation decreases with increasing rho
```{r}
# Create data frame for plot
y <- rbind(rho0, rho4, rho10)
df <- data_frame(tree_height1 = y[,1], tree_height2 = y[,2], rho = as.character(y[,3]))
# Plot
plot1 <- ggplot(df, aes(x = tree_height1, y = tree_height2, color = rho)) + 
  geom_point(aes(color = rho), size = 3, alpha = 0.1) + 
  geom_point(aes(color = rho),shape = 1, size = 3, stroke = 0.2) + 
  scale_y_continuous(name = "Tree height locus 1", 
                      limits = c(0,10), expand = c(0, 0)) + 
  scale_x_continuous(name = "Tree height locus 2", 
                      limits = c(0,10),expand = c(0, 0)) +
  scale_color_manual(values=c("#999999", "#E69F00", "#56B4E9"))+
  theme_pubr() +
  theme(legend.position = "bottom", legend.direction = "horizontal")
# Density functions of tree heights in locus 1
dens1 <- ggplot(df, aes(x = tree_height1, color = rho)) + 
  geom_density(alpha = 0.4) + 
  theme_void() + 
  scale_color_manual(values=c("#999999", "#E69F00", "#56B4E9"))+
  theme(legend.position = "none")

# Density functions of tree heights in locus 2
dens2 <- ggplot(df, aes(x = tree_height2, color = rho)) + 
  geom_density(alpha = 0.4) + 
  theme_void() + 
  theme(legend.position = "none") + 
  scale_color_manual(values=c("#999999", "#E69F00", "#56B4E9"))+
  coord_flip()
# Final plot
dens1 + plot_spacer() + plot1 + dens2 + 
  plot_layout(ncol = 2, nrow = 2, widths = c(5, 2), heights = c(2, 5))
```

# CHECK: Plot expected density of the tree heights against observed density for rho = 0
  Observation: density curve drops when very close to 0, because of the density package
```{r}
# Create data frame for plot
x <- seq(0, 6, 0.1)
rho0_expected <- data_frame(treeheight = (cbind(dMPH(x, MPH_gen0)[1,])))
rho0_expected['x'] <-x
df <- data_frame(tree_height1 = rho0[,1], tree_height2 = rho0[,2], )

# Plot
plot1 <- ggplot(df, aes(x = tree_height1, y = tree_height2))+ 
  geom_point(size = 3, alpha = 0.03, color = "#E69F00") + 
  scale_y_continuous(name = "Tree height locus 1", 
                      limits = c(0,6), expand = c(0, 0)) + 
  scale_x_continuous(name = "Tree height locus 2", 
                      limits = c(0,6),expand = c(0, 0)) +
  theme_pubr()

# Density function of tree height in locus 1
dens1 <- ggplot(df) + 
  geom_density(aes(x = tree_height1),alpha = 0.4, color = "#E69F00") + 
  geom_line(data = rho0_expected, aes(x = x, y = treeheight), color="#999999")+
  theme_void()+  
  theme(legend.position = "none")
# Density function of tree height in locus 2
dens2 <- ggplot(df) + 
  geom_density(aes(x = tree_height2, color = "Simulated"),alpha = 0.4) + 
  geom_line(data = rho0_expected, aes(x = x, y = treeheight, color="Expected"))+
  theme_void() +
  scale_color_manual(name='Density',
                     breaks=c('Simulated', 'Expected'),
                     values=c('Simulated'='#E69F00', 'Expected'='#999999'))+
  theme(legend.position = "right")+
  coord_flip()
# Final plot
dens1 + plot_spacer() + plot1 + dens2 + 
  plot_layout(ncol = 2, nrow = 2, widths = c(5, 2), heights = c(2, 5))
```

# CHECK: Compare density of tree heights for standard coalescent model and implemented coalescent with recombination model 
  Idea: For rho = 0 the returned variables should be the same as the once from 
  the standard coalescent model.
```{r}
# Initialize
n <- 5
rho <- 0
sModel <- basicCoalModel(n)
MPH_simple <- MPH(sModel[["subIntMat"]], sModel[["initProbs"]], sModel[["rewardMatTotal"]])

# Create MPH for advanced model with rho = 0
test_gen <- IntensityMatrix(n,rho)
MPH_gen <- MPH(test_gen[["subIntMat"]], test_gen[["initProbs"]], test_gen[["rewardMatTotal"]])

# Plot density function of standard coalescent model
x <- seq(0, 10, 0.1)
y <- dMPH(x, MPH_simple)
plot(x, y[1,], type = 'p', col = 'orange', ylab = "Density", xlab = "Tree height")
# Add to plot density function of implemented model  
y2 <- dMPH(x, MPH_gen)
lines(x, y2[1,], col = 'blue',  type = 'l')
legend(6, 0.5, legend=c(expression('Simple, T'[total]), expression('generated, T'[total])),
       col=c("orange", "blue"), lty=1)
title(paste('Density function (n=',n,', rho=0)'))
```

# CHECK: Compare covarianve equations from Wakeley's book to covariance generated by Phase-type distributions (n=2)
```{r}
#create data frame
df <- data.frame(matrix(ncol = 4, nrow = 0))
colnames(df) <- c('Recombination.Rate', 'Generated', 'Wakeley', 'Start.state')

#generate cov-values for rho in 0-10
for(rho in seq(0, 10, by=0.5)){
      #generate rewards, S and a
  Output <- IntensityMatrix(2, rho)
      # create MPH
      MPH_gen_1 <- MPH(Output[["subIntMat"]], Output[["initProbs"]], Output[["rewardMatMRCA"]])
      MPH_gen_2 <- MPH(Output[["subIntMat"]], c(0,1,0,0,0,0,0), Output[["rewardMatMRCA"]])
      MPH_gen_3 <- MPH(Output[["subIntMat"]], c(0,0,0,0,1,0,0), Output[["rewardMatMRCA"]])
      #append df with rho and cov
      df[nrow(df) + 1,] = c(rho, var(MPH_gen_1)[1,2], cov_1(rho), 1)
      df[nrow(df) + 1,] = c(rho, var(MPH_gen_2)[1,2], cov_2(rho), 2)
      df[nrow(df) + 1,] = c(rho, var(MPH_gen_3)[1,2], cov_3(rho), 3)

}
# create line plot 
# melt data frame into long format
df<- melt(df ,  id.vars = c('Recombination.Rate', 'Start.state'), variable.name = 'Covariance')
# plot 
ggplot(df, aes(x = Recombination.Rate, y = value, group = interaction(Start.state,Covariance) , colour = Covariance, linetype = Covariance )) +
  geom_line()+ylab("Covariance") +xlab(expression(rho))+scale_linetype_manual(values=c("solid", "dotted"))+
  ggtitle("Compare covarianve equation (Wakeley) \nto covariance generated by Phase-type distributions (n=2)\nfor X in {1,2,3}")+ scale_color_manual(values=c("orange", "blue"))

```

# Heatmap: Possible transitions between states
  # red: coalescence event, blue: recombination event
```{r}
n <- 4
# generate matrixes for recombination and mutation separatly to add categories
Output_rec <- IntensityMatrix(n, 1)
Output_mut <- IntensityMatrix(n, 0)

data_rec <- as.data.frame(Output_rec[["subIntMat"]])%>%
  rownames_to_column("Y") %>%
  pivot_longer(-c(Y), names_to = "X", values_to = "recombination")

data_mut <- as.data.frame(Output_mut[["subIntMat"]])%>%
  rownames_to_column("Y") %>%
  pivot_longer(-c(Y), names_to = "X", values_to = "coalescence")

#join two tables and assign type: coalescence, recombination, exit or nothing
joined <- inner_join(data_rec, data_mut, by = c("X","Y")) %>% mutate(total = (coalescence + recombination)) %>% mutate(type = case_when(coalescence > 0 ~ "Coal", recombination > 0 ~ "Rec", total < 0 ~ "Neg", total == 0 ~ "0")) 

joined$X <- with(joined,factor(X,levels = rev(sort(unique(X)))))
#table with rewards
data_rec$Y
reward_table <- as.data.frame(Output_mut[["rewardMatTotal"]])
row.names(reward_table) <- rownames(Output_mut[["subIntMat"]])
colnames(reward_table) <- c("R1","R2")
reward_table_plot <-tableGrob(reward_table, theme = ttheme_minimal())
#to fit table on pdf next to plot
reward_table_plot$heights <- unit(rep(1/nrow(reward_table_plot), nrow(reward_table_plot)), "npc")

heatmap_plot <-joined %>% ggplot(aes(X, y = Y)) + geom_tile(aes(fill = type))+scale_fill_manual(values = c("Rec" = "blue", "Coal" = "red", "Neg" = "grey", "0" = "white"))+theme(axis.text.x = element_text(angle=90, vjust=.5, hjust=1)) +theme(
      axis.title.x=element_blank(),
      axis.title.y=element_blank())
if (n > 4){
  heatmap_plot <- heatmap_plot + theme(
      axis.text.x=element_blank(),
      axis.text.y=element_blank())
}

#plot heatmap
heatmap_plot +coord_fixed()

# save plot
#png(paste("Slide17_", n, ".png"))
#gtable_combine(heatmap_plot + reward_table_plot, along=1)
#dev.off() 

```

# Represent Markov process as graph 
```{r}
Output <- IntensityMatrix(2, 2)
# create MPH
Y_network <- phase_type_to_network(PH(Output[["subIntMat"]]))
set.seed(200)
plot(Y_network, vertex.size = 20,layout = layout_with_fr(Y_network,  weights = rep(1, length(E(Y_network)))),asp = .66, rescale = T)
#vertex.shape = "square"
```

# Figure 6.1:
The size of the state space as a function of the sample size n 
```{r}
eq <- function(x){x*((2*x**2 + 9*x + 1)/6 )}
state_df <- data.frame("n" = seq(0,100,1))
state_df <- state_df %>% mutate(states = n*((2*n**2 + 9*n + 1)/6))
p<-ggplot(state_df,aes(x = n,y = states))+ geom_point(colour = "#999999", size = 0.5) + ylab("# states")
#ggsave(file="nuber_states.pdf", plot=p, width=5, height=3)
```